Final Project: 2011 Dallas Mavericks
Dirk top 3 PF OAT

NBA Data
```{r}
nba = read.csv("nba_data.csv", header = T)
nba$MoV = nba$home_score - nba$away_score
nba11 = nba[nba$season == 2010,]
nbastop11 = nba[nba$season <= 2010,]
```


4 Factors
```{r}
szn = read.csv("nba_2011_data.csv", header = T)
szn$eFG = (szn$FG + .5*szn$X3P) / szn$FGA
szn$ORPct = szn$ORB / (szn$ORB + szn$DRB)
szn$DRPct = szn$DRB / (szn$DRB + szn$ORB)
szn$TOR = szn$TOV / (szn$FGA + 0.44*szn$FTA + szn$TOV)
szn$FTR = szn$FTA / szn$FGA

#effective Field Goal
mean(szn$eFG)
sd(szn$eFG)
boxplot(szn$eFG, horizontal = T, xlab = "effective Field Goal %")
stripchart(szn$eFG, method = "jitter", pch = 19, col = "red", add = T)
stripchart(szn$eFG[szn$Team == "Dallas Mavericks*"], method = "jitter", pch = 19, col = "blue", cex = 3, add = T)

#Rebounding
mean(szn$ORPct)
sd(szn$ORPct)
boxplot(szn$ORPct, horizontal = T, xlab = "Offensive Rebound %")
stripchart(szn$ORPct, method = "jitter", pch = 19, col = "red", add = T)
stripchart(szn$ORPct[szn$Team == "Dallas Mavericks*"], method = "jitter", pch = 19, col = "blue", cex = 3, add = T)

mean(szn$DRPct)
sd(szn$DRPct)
boxplot(szn$DRPct, horizontal = T, xlab = "Defensive Rebound %")
stripchart(szn$DRPct, method = "jitter", pch = 19, col = "red", add = T)
stripchart(szn$DRPct[szn$Team == "Dallas Mavericks*"], method = "jitter", pch = 19, col = "blue", cex = 3, add = T)

#Turnovers
mean(szn$TOR)
sd(szn$TOR)
boxplot(szn$TOR, horizontal = T, xlab = "Turnover %")
stripchart(szn$TOR, method = "jitter", pch = 19, col = "red", add = T)
stripchart(szn$TOR[szn$Team == "Dallas Mavericks*"], method = "jitter", pch = 19, col = "blue", cex = 3, add = T)

#Free Throw Rate
mean(szn$FTR)
sd(szn$FTR)
boxplot(szn$FTR, horizontal = T, xlab = "Free Throw Rate ")
stripchart(szn$FTR, method = "jitter", pch = 19, col = "red", add = T)
stripchart(szn$FTR[szn$Team == "Dallas Mavericks*"], method = "jitter", pch = 19, col = "blue", cex = 3, add = T)

```

NBA Scoring Dist
```{r}
#charts
hist(nba11$home_score, breaks = 30, main = "Home Scoring", xlab = "Home Score")
abline(v = mean(nba11$home_score[nba11$home_team == "Dallas Mavericks"]), col = "blue")
abline(v = mean(nba11$home_score), col = "red")

hist(nba11$away_score, breaks = 30, main = "Away Scoring", xlab = "Away Score")
abline(v = mean(nba11$away_score[nba11$away_team == "Dallas Mavericks"]), col = "blue")
abline(v = mean(nba11$away_score), col = "red")

hist(nba11$MoV, breaks = 20, main = "Margin of Victory", xlab = "Margin of Victory")
abline(v = mean(nba11$MoV), col = "red")

plot(nba11$home_score ~ nba11$away_score, col = ifelse(nba11$MoV > 0, "green3", "red"))

#numeric
cor(nba11$home_score, nba11$away_score)
mean(nba11$home_score)
mean(nba11$away_score)
```



Pythagorean Win Percentage - setting up data
```{r}
nba11 = nba11[nba11$game_type == "r",]
Team = unique(nba11$home_team)
nbaTeams = as.data.frame(Team)

for (team in nbaTeams$Team){
  home_games = subset(nba11, home_team == team)
  away_games = subset(nba11, away_team == team)
  
  nbaTeams$wins[nbaTeams$Team == team] = sum(home_games$MoV > 0) + sum(away_games$MoV < 0) #wins 
  nbaTeams$losses[nbaTeams$Team == team] = sum(home_games$MoV < 0) + sum(away_games$MoV > 0) #losses 
  
  nbaTeams$pointsFor[nbaTeams$Team == team] = (sum(home_games$home_score) + sum(away_games$away_score))
  nbaTeams$pointsAgainst[nbaTeams$Team == team] = (sum(home_games$away_score) + sum(away_games$home_score))
  
}
nbaTeams$wpct = nbaTeams$wins / (nbaTeams$wins + nbaTeams$losses)
```

Pythagorean Win Percentage - Calculation
```{r}
k = 14.112

model = lm(data = nbaTeams,  (log(wins/losses)) ~ (log(pointsFor/pointsAgainst)))
summary(model)
nbaTeams$pythag_win_pcent = (nbaTeams$pointsFor^k) / (nbaTeams$pointsFor^k + nbaTeams$pointsAgainst^k)
nbaTeams$error = nbaTeams$wpct - nbaTeams$pythag_win_pcent
cor = cor(nbaTeams$wpct, nbaTeams$pythag_win_pcent)
```

Pythagorean Win Percentage - Graphs
```{r}
plot(nbaTeams$wpct ~ nbaTeams$pythag_win_pcent, xlab = "Pythagorean Win Percentage", ylab = "Real Win Percentage", main = "Pythagorean vs Real %")
abline(a = 0, b = 1, col = "red")
abline(a = 0, b = cor, col = "blue", lty = 2)

plot(nbaTeams$error ~ nbaTeams$pythag_win_pcent, xlab = "Pythagorean Win Percentage", ylab = "Difference from Real Win Percentage (Error)", main = "Distribution of Error")
abline(h = 0, col = "red")
```



Bradley Terry - setting up data
```{r}
nba_data = as.data.frame(matrix(nrow = nrow(nba11), ncol = 32))
nba_data[is.na(nba_data)] = 0
teams = unique(nba11$home_team)
colnames(nba_data) = c("MoV", "Loc", teams)

for(i in 1:length(teams)){
  home = as.integer(nba11$home_team == teams[i])
  away = -1 * as.integer(nba11$away_team == teams[i])
  team_locations = home+away
  nba_data[,i+2] = team_locations
}

nba_data$MoV = nba11$MoV
nba_data$Loc = nba11$location

bt_nba_model = lm(MoV ~ 0 + ., data = nba_data)
summary(bt_nba_model)
```

Bradley Terry Adjustments
```{r}
coefficients = coef(bt_nba_model, complete = TRUE)
coefficients[length(coefficients)] = 0

adjustment = -1*sum(coefficients[2:length(coefficients)]) / (length(coefficients) - 1)
ratings = c(coefficients[1], coefficients[2:length(coefficients)] + adjustment)  
ratings
```

Bradley Terry Graphics
```{r}
table = data.frame(Team = names(ratings), Rating = as.numeric(ratings))
table$Coef = bt_nba_model$coefficients
table = table[, c("Team", "Coef", "Rating")]
```



Elo Ratings - Setup + Calc
```{r}
elos = read.csv("nba_initial_elos.csv", header = T)
scores = nbastop11

weight = 8.5 
hfa = 77.47

if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}

team = "Dallas Mavericks"
first_season = 2010
last_season = 2010
team_results = data.frame(matrix(ncol = 10, nrow = 0))
colnames(team_results) = c("opponent", "pregame_elo", "opponent_pregame_elo", "win_probability", "result", "team_score", "opponent_score", "elo_adjustment", "postgame_elo", "opponent_postgame_elo")

# Iterate through all games in the sport's history
for(i in 1:nrow(scores)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(elos$team == scores$home_team[i])
    away_index = which(elos$team == scores$away_team[i])
    
    
    # Find home and away team Elo ratings
    home_elo = elos$rating[home_index]
    away_elo = elos$rating[away_index]
    
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*scores$location[i]))/400) + 1)
    
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(scores$home_score[i] - scores$away_score[i])   
    
    
    # Determine home team result
    if(scores$home_score[i] > scores$away_score[i]) { 
        home_result = 1  # Home team wins
    } else if(scores$home_score[i] < scores$away_score[i]) { 
        home_result = 0  # Home team loses
    } else { 
        home_result = 0.5  # Tie
    }
  
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
    
    
    # Adjust Elo ratings - add point to winner and subtract points from loser
    elos$rating[home_index] = elos$rating[home_index] + home_elo_adjustment
    elos$rating[away_index] = elos$rating[away_index] - home_elo_adjustment

    
    # Add game information to team result data frame for each team game of the team specified above if team and season both match
    if(scores$season[i] >= first_season & scores$season[i] <= last_season & (scores$home_team[i] == team | scores$away_team[i] == team)) {
        if(scores$home_team[i] == team) {  # If specified team was at home
            team_results[nrow(team_results) + 1,] = c(scores$away_team[i], elos$rating[home_index] - home_elo_adjustment, elos$rating[away_index] + home_elo_adjustment, win_prob, home_result, scores$home_score[i], scores$away_score[i], home_elo_adjustment, elos$rating[home_index], elos$rating[away_index])
        } else {  # If specified team was away
            team_results[nrow(team_results) + 1,] = c(scores$home_team[i], elos$rating[away_index] + home_elo_adjustment, elos$rating[home_index] - home_elo_adjustment, 1-win_prob, 1-home_result, scores$away_score[i], scores$home_score[i], -1*home_elo_adjustment, elos$rating[away_index], elos$rating[home_index])
        }
    }
    
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {  # New season
        for(j in 1:nrow(elos)) {  # For each team
            if(scores$season[i] >= elos$inaugural_season[j]) {  # Check if team existed
                # Move each team's Elo rating back towards 1500 by 1/3 of the difference
                elos$rating[j] = elos$rating[j] - (elos$rating[j] - 1500)/3
            }
        }
        
        
        # Identify all teams that existed at beginning of following season
        existing_teams = elos[which(elos$inaugural_season <= (scores$season[i] + 1)),]
        
        
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
        
        
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(elos)) {  # For each team
            if((scores$season[i] + 1) >= elos$inaugural_season[j]) {  # Check if team existed
                elos$rating[j] = elos$rating[j] + expansion_adjustment  # Update ratings if so
            }
        }
    }
    if (scores$season[i] == 2009) {
          elos$beginning_ratings_2010 = elos$rating
        }
}

if (first_season == last_season) {
    team_name = paste(first_season, team)
} else {
    team_name = paste(paste(first_season, last_season, sep = "-"), team)
}

plot(team_results$postgame_elo, type = "l", xlab = team_name, ylab = "Elo Rating")
abline(h = 1500)
```

Elo Ratings - Graphics
```{r}
elotable = data.frame(elos[ !is.na(elos$conference),])
elotable = elotable[, c("team", "conference", "division", "beginning_ratings_2010","rating")]
elotable$change = elotable$rating-elotable$beginning_ratings_2010
```



Monte Carlo - Setup
```{r}
scores = read.csv("nba_data.csv", header = T)
team_info = read.csv("nba_initial_elos.csv", header = T)
```

Monte Carlo - Calc
```{r}

simulated_season = 2010

conferences = na.omit(unique(team_info$conference))
divisions = na.omit(unique(team_info$division))

pre_season = scores[which(scores$season < simulated_season & scores$season >= 1901),]

season_schedule = scores[which(scores$season == simulated_season & (scores$game_type == "r" | scores$game_type == "np")),]



# Input the optimal k factors (weight) and home field advantage
hfa = 77.47
weight = 8.5

# Identify if ties exist in the sport, which determines if an additional 0.5 needs to be added to the margin of victory
if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}

# Iterate through all games in the sport's history up to season being simulated
for(i in 1:nrow(pre_season)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(team_info$team == pre_season$home_team[i])
    away_index = which(team_info$team == pre_season$away_team[i])
    
    # Find home and away team Elo ratings
    home_elo = team_info$rating[home_index]
    away_elo = team_info$rating[away_index]
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*pre_season$location[i]))/400) + 1)
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(pre_season$home_score[i] - pre_season$away_score[i])
    
    # Determine home team result
    if(pre_season$home_score[i] > pre_season$away_score[i]) { 
    home_result = 1  # Home team wins
    } else if(pre_season$home_score[i] < pre_season$away_score[i]) { 
    home_result = 0  # Home team loses
    } else { 
    home_result = 0.5  # Tie
    }
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)

  
    # Adjust Elo ratings - add point to winner and subtract points from loser
    team_info$rating[home_index] = team_info$rating[home_index] + home_elo_adjustment
    team_info$rating[away_index] = team_info$rating[away_index] - home_elo_adjustment
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {
        for(j in 1:nrow(team_info)) {
            if(scores$season[i] >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] - (team_info$rating[j] - 1500)/3
                }
        }
    
        # Identify all teams that existed at beginning of following season
        existing_teams = team_info[which(team_info$inaugural_season <= (scores$season[i] + 1)),]
    
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
    
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(team_info)) {
            if((scores$season[i] + 1) >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] + expansion_adjustment
            }
        }
    }
}

# Set seed for replication purposes
set.seed(37)

# Determine number of times to simulate the season
iterations = 1000

# Create data frame to hold Elo ratings, actual win totals, and simulation results
results = data.frame(matrix(0, ncol = 6, nrow = nrow(team_info)))
colnames(results) = c("team", "starting_elo", "ending_elo", "actual_wins", "average_wins", "division_titles")
results$team = team_info$team

# Create data frame to hold number of wins by each team in each iteration
win_totals = data.frame(matrix(0, ncol = nrow(team_info), nrow = iterations))
colnames(win_totals) = team_info$team

# Simulate the season the given number of times
for(i in 1:iterations) {
    if(i %% 1000 == 0) {print(i)}
    season_stats = team_info[,which(colnames(team_info) != "inaugural_season")]
    season_stats$wins = 0
    season_stats$rand = runif(nrow(team_info))
    
    # Simulate each game in current season
    for(j in 1:nrow(season_schedule)) {
        # Find indices corresponding to home and away teams for current game
        home_index = which(season_stats$team == season_schedule$home_team[j])
        away_index = which(season_stats$team == season_schedule$away_team[j])
        
        # Find home and away team Elo ratings
        home_elo = season_stats$rating[home_index]
        away_elo = season_stats$rating[away_index]
        
        # Calculate home team win probability
        win_prob = 1 / (10^((away_elo - (home_elo + hfa*season_schedule$location[j]))/400) + 1)
        u = runif(1)  # Generate a random number used to determine the winner of the game
        
        # Determine which team wins the simulated game and increment their win total by 1
        if(u < win_prob) {
            season_stats$wins[home_index] = season_stats$wins[home_index] + 1
        } else {
            season_stats$wins[away_index] = season_stats$wins[away_index] + 1
        }
    
        # Calculate actual margin of victory - must be positive
        score_diff = abs(season_schedule$home_score[j] - season_schedule$away_score[j])
        
        # Determine home team result
        if(season_schedule$home_score[j] > season_schedule$away_score[j]) { 
          home_result = 1  # Home team wins
        } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) { 
          home_result = 0  # Home team loses
        } else { 
          home_result = 0.5  # Tie
        }
        
        # Calculate amount each team's Elo rating is adjusted by
        home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
        
        # Adjust Elo ratings after game has been simulated to get team's new strength
        season_stats$rating[home_index] = season_stats$rating[home_index] + home_elo_adjustment
        season_stats$rating[away_index] = season_stats$rating[away_index] - home_elo_adjustment
    
        # Track season wins
        if(i == 1) {
            if(season_schedule$home_score[j] > season_schedule$away_score[j]) {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 1
            } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) {
                results$actual_wins[away_index] = results$actual_wins[away_index] + 1
            } else {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 0.5
                results$actual_wins[away_index] = results$actual_wins[away_index] + 0.5
            }
        }
    }

    # Define data frame that contains division winners
    division_winners = data.frame(matrix(ncol = 6, nrow = 0))
    colnames(division_winners) = c("team", "conference", "division", "rating", "wins", "rand")
  
    # For each division
    for(div in divisions) {
        div_standings = season_stats[which(season_stats$division == div),]  # Identify all teams in current division
        div_standings = div_standings[order(-div_standings$wins, -div_standings$rand),]  # Sort division by wins and random number
        division_winners = rbind(division_winners, div_standings[1,])  # Add division winner to 'division_winners' data frame
    }
  
    # Save end of season win totals (from column 5 of season_stats) for each iteration in the win_totals data frame
    win_totals[i,] = t(season_stats[5])
    
    # Add number of wins for each team during this iteration to sum
    results$average_wins = results$average_wins + season_stats$wins
  
    # Increment the number of division titles for each division winner by 1
    for(team in division_winners$team) {
        index = which(season_stats$team == team)  # Index of division winner
        results$division_titles[index] = results$division_titles[index] + 1  # Increment division titles
    }
}
 
# Calculate average number of wins across all iterations
results$starting_elo = team_info$rating
results$ending_elo = season_stats$rating
results$average_wins = results$average_wins/iterations
results$division_titles = results$division_titles/iterations
results
```

Monte Carlo - Graphics
```{r}
results = results[results$actual_wins != 0,]
results$error = results$actual_wins - results$average_wins

hist(results$actual_wins-results$average_wins, xlab = "Actual - Average wins", main = "Histogram of Residuals")

color_logic = ifelse(results$team == "Dallas Mavericks", "blue", ifelse(results$team == "Oklahoma City Thunder", "orange", ifelse(results$team == "Toronto Raptors", "black", ifelse(results$actual_wins > results$average_wins, "green", "red"))))

plot(results$actual_wins ~ results$average_wins, pch = 19, col = color_logic, xlab = "Average Wins", ylab = "Actual Wins", main = "Actual vs Average Plot")

```


Overachievers
```{r}
hist(win_totals$`Oklahoma City Thunder`, xlab = "Games Won", main = "OKC Win Total Histogram")
abline(v = results$actual_wins[results$team == "Oklahoma City Thunder"], col = "orange")

hist(win_totals$`Toronto Raptors`, xlab = "Games Won", main = "TOR Win Total Histogram")
abline(v = results$actual_wins[results$team == "Toronto Raptors"], col = "red")
```









